Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2023-05-15 doc/manuscript/ESSDmanuscript.Rmd
refs
#39

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC]. 
While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although thhese do tend to be censused when they consitute a large proportion of the biomass [@refs_from_ForC].
Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

@@

New data collection and analysis is needed to fill notable knowledge gaps.
While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021], production of an accurate global map of forest C stocks remains an ongoing challenge [@refs; @labrière_forest_2023]. 
Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@refs].
Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@refs].

@@

Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
Given observations of increasing tree mortality in some forested regions [@refs], including through severe natural disturbance [@refs], better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@refs].
Significant investment in ground-based forest research will be critical to filling these gaps.

@@

Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are compatible with IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest C accounting. 
This challenge is heightened by the fact that forests are changing rapidly [@refs], and data collected a decade or more in the past may no longer be relevant.
This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
We view this publication as one step towards that goal.

>>>>

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_good_2003;@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" are not provided*<!--verify (issue 40)-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@lutz_global_2018; @piponiot_distribution_2022; @refs_from_ForC], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs_from_ForC]. 
While IPCC guidance specifies that all living vegetation should be included in biomass estimates, forest censuses and biomass estimates do not consistently include life forms other than dicot trees (e.g., lianas, ferns, palms, bamboo), although thhese do tend to be censused when they consitute a large proportion of the biomass [@refs_from_ForC].
Further, it is important to note that the IPCC definition of aboveground biomass excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

@@

New data collection and analysis is needed to fill notable knowledge gaps.
While aboveground biomass stocks in particular have received -- and continue to receive -- by far the most research attention [Table `r table_variables`, @anderson-teixeira_carbon_2021;@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], production of an accurate global map of forest C stocks remains an ongoing challenge [@araza_decade_2023]. 
Other pools and variables remain very poorly quantified [Table `r table_variables`, @anderson-teixeira_carbon_2021], introducing substantive uncertainties into global forest C budgets [@pan_large_2011;@harris_global_2021].
Furthermore, data distribution is uneven across forest types and geographical regions (Figs. `r fig_map`-`r fig_histograms`).
For instance, data on C cycling of tropical forests -- particularly in Africa -- remains relatively sparse, in large part due to substantial barriers to data collection and distribution [@delima_making_2022].
Significant investment in research and researchers focused on ground-based measurement of forest C in such regions will be important to filling knowledge gaps in forest C cycling [@delima_making_2022;@araza_decade_2023; @labrière_forest_2023].

@@

Filling knowledge gaps in other EFDB-relevant variables will require more effort, but this effort is warranted given their importance for forest C accounting.
Although aboveground biomass is the most studied variable considered here (Table `r table_variables`) and is the target of satellite missions [@dubayah_global_2020; @quegan_european_2019; @nisar_nasaisro_2018], significant ground-based research effort is required to create accurate global maps of forest biomass and changes therein [@duncanson_importance_2019; @labrière_forest_2023].
Given observations of increasing tree mortality in some forested regions [@mcdowell_pervasive_2020], better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite substantial and highly uncertain in many parts of the world [@tifafi_large_2018].
Significant investment in ground-based forest research will be critical to filling these gaps.

@@

Going forward, forest researchers can make their research more useful for forest C accounting under IPCC guidelines by calculating and reporting results in ways that are compatible with IPCC guidelines (Tables `r table_pools`, `r table_recommendations`).
In addition, substantial investments in research and researchers focused on ground-based measurement of forest C will be required to fill knowledge gaps and thereby increase the accuracy of forest C accounting. 
This challenge is heightened by the fact that forests are changing rapidly [e.g., @mcdowell_pervasive_2020], and data collected a decade or more in the past may no longer be relevant.
This heightens the need for an efficient system of making forest C data accessible for national greenhouse gas inventories.
We view this publication as one step towards that goal.

### 2023-05-15 doc/manuscript/ESSDmanuscript.tex
refs
#39

Aboveground biomass, which is typically \textless200 t C
ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
biomass of living vegetation above the soil including stems, stumps,
branches, bark, seeds, and foliage''
\citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
understory may be excluded the understory if it constitutes a ``minor''
component, \emph{where quantitative definitions of ``understory'' and
``minor'' are not provided}, but where a commonly applied minimum size
sampling threshold for mature forests would be 10 cm stem diameter at
breast height (DBH). A recent study characterizing the contributions of
trees in different DBH classes to ecosystem C stocks and fluxes found
that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
aboveground biomass, \textasciitilde17\% aboveground woody net primary
productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
savannas, small trees and shrubs contribute a much larger proportion of
C stocks and fluxes \citep{piponiot_distribution_2022, refs_from_ForC},
and, correspondingly, biomass estimates for these ecosystems tend
include smaller size classes \citep[e.g.,][]{refs_from_ForC}. While IPCC
guidance specifies that all living vegetation should be included in
biomass estimates, forest censuses and biomass estimates do not
consistently include life forms other than dicot trees (e.g., lianas,
ferns, palms, bamboo), although thhese do tend to be censused when they
consitute a large proportion of the biomass \citep{refs_from_ForC}.
Further, it is important to note that the IPCC definition of aboveground
biomass excludes standing dead wood, which is included in remote sensing
biomass estimates \citep{duncanson_aboveground_2021}.

@@

New data collection and analysis is needed to fill notable knowledge
gaps. While aboveground biomass stocks in particular have received --
and continue to receive -- by far the most research attention
\citep[Table 2,][]{anderson-teixeira_carbon_2021}, production of an
accurate global map of forest C stocks remains an ongoing challenge
\citep{refs, labrière_forest_2023}. Other pools and variables remain
very poorly quantified \citep[Table
2,][]{anderson-teixeira_carbon_2021}, introducing substantive
uncertainties into global forest C budgets \citep{refs}. Furthermore,
data distribution is uneven across forest types and geographical regions
(Figs. 2-3). For instance, data on C cycling of tropical forests --
particularly in Africa -- remains relatively sparse, in large part due
to substantial barriers to data collection and distribution
\citep{delima_making_2022}. Significant investment in research and
researchers focused on ground-based measurement of forest C in such
regions will be important to filling knowledge gaps in forest C cycling
\citep{delima_making_2022, refs}.

@@

Filling knowledge gaps in other EFDB-relevant variables will require
more effort, but this effort is warranted given their importance for
forest C accounting. Although aboveground biomass is the most studied
variable considered here (Table 2) and is the target of satellite
missions
\citep{dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
significant ground-based research effort is required to create accurate
global maps of forest biomass and changes therein
\citep{duncanson_importance_2019, labrière_forest_2023}. Given
observations of increasing tree mortality in some forested regions
\citep{refs}, including through severe natural disturbance \citep{refs},
better characterization of forest dead wood will be critical.
Additionally, C stocks in forest organic horizons and soils can be quite
substantial and highly uncertain in many parts of the world
\citep{refs}. Significant investment in ground-based forest research
will be critical to filling these gaps.

@@

Going forward, forest researchers can make their research more useful
for forest C accounting under IPCC guidelines by calculating and
reporting results in ways that are compatible with IPCC guidelines
(Tables 1, 3). In addition, substantial investments in research and
researchers focused on ground-based measurement of forest C will be
required to fill knowledge gaps and thereby increase the accuracy of
forest C accounting. This challenge is heightened by the fact that
forests are changing rapidly \citep{refs}, and data collected a decade
or more in the past may no longer be relevant. This heightens the need
for an efficient system of making forest C data accessible for national
greenhouse gas inventories. We view this publication as one step towards
that goal.

>>>>

Aboveground biomass, which is typically \textless200 t C
ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
biomass of living vegetation above the soil including stems, stumps,
branches, bark, seeds, and foliage''
\citep{ipcc_good_2003, ipcc_2006_2006}. IPCC's guidance is that the
understory may be excluded the understory if it constitutes a ``minor''
component, \emph{where quantitative definitions of ``understory'' and
``minor'' are not provided}, but where a commonly applied minimum size
sampling threshold for mature forests would be 10 cm stem diameter at
breast height (DBH). A recent study characterizing the contributions of
trees in different DBH classes to ecosystem C stocks and fluxes found
that trees 1 - 10 cm DBH contributed up to \textasciitilde8\%
aboveground biomass, \textasciitilde17\% aboveground woody net primary
productivity (\(ANPP_{woody.stem}\)), and \textasciitilde20\% woody
mortality (\(M_{woody}\)) of mature closed-canopy forests worldwide
\citep{piponiot_distribution_2022}. In regrowth forests, woodlands, or
savannas, small trees and shrubs contribute a much larger proportion of
C stocks and fluxes
\citep{lutz_global_2018, piponiot_distribution_2022, refs_from_ForC},
and, correspondingly, biomass estimates for these ecosystems tend
include smaller size classes \citep[e.g.,][]{refs_from_ForC}. While IPCC
guidance specifies that all living vegetation should be included in
biomass estimates, forest censuses and biomass estimates do not
consistently include life forms other than dicot trees (e.g., lianas,
ferns, palms, bamboo), although thhese do tend to be censused when they
consitute a large proportion of the biomass \citep{refs_from_ForC}.
Further, it is important to note that the IPCC definition of aboveground
biomass excludes standing dead wood, which is included in remote sensing
biomass estimates \citep{duncanson_aboveground_2021}.

@@

New data collection and analysis is needed to fill notable knowledge
gaps. While aboveground biomass stocks in particular have received --
and continue to receive -- by far the most research attention
\citep[Table
2,][]{anderson-teixeira_carbon_2021, dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
production of an accurate global map of forest C stocks remains an
ongoing challenge \citep{araza_decade_2023}. Other pools and variables
remain very poorly quantified \citep[Table
2,][]{anderson-teixeira_carbon_2021}, introducing substantive
uncertainties into global forest C budgets
\citep{pan_large_2011, harris_global_2021}. Furthermore, data
distribution is uneven across forest types and geographical regions
(Figs. 2-3). For instance, data on C cycling of tropical forests --
particularly in Africa -- remains relatively sparse, in large part due
to substantial barriers to data collection and distribution
\citep{delima_making_2022}. Significant investment in research and
researchers focused on ground-based measurement of forest C in such
regions will be important to filling knowledge gaps in forest C cycling
\citep{delima_making_2022, araza_decade_2023, labrière_forest_2023}.

@@

Filling knowledge gaps in other EFDB-relevant variables will require
more effort, but this effort is warranted given their importance for
forest C accounting. Although aboveground biomass is the most studied
variable considered here (Table 2) and is the target of satellite
missions
\citep{dubayah_global_2020, quegan_european_2019, nisar_nasaisro_2018},
significant ground-based research effort is required to create accurate
global maps of forest biomass and changes therein
\citep{duncanson_importance_2019, labrière_forest_2023}. Given
observations of increasing tree mortality in some forested regions
\citep{mcdowell_pervasive_2020}, better characterization of forest dead
wood will be critical. Additionally, C stocks in forest organic horizons
and soils can be quite substantial and highly uncertain in many parts of
the world \citep{tifafi_large_2018}. Significant investment in
ground-based forest research will be critical to filling these gaps.

@@

Going forward, forest researchers can make their research more useful
for forest C accounting under IPCC guidelines by calculating and
reporting results in ways that are compatible with IPCC guidelines
(Tables 1, 3). In addition, substantial investments in research and
researchers focused on ground-based measurement of forest C will be
required to fill knowledge gaps and thereby increase the accuracy of
forest C accounting. This challenge is heightened by the fact that
forests are changing rapidly \citep[e.g.,][]{mcdowell_pervasive_2020},
and data collected a decade or more in the past may no longer be
relevant. This heightens the need for an efficient system of making
forest C data accessible for national greenhouse gas inventories. We
view this publication as one step towards that goal.
